---
title: "DSC 011 S26 — Lecture 32 Demo"
subtitle: "UNIX Text Utilities, III, Compiling"
author: "FirstName LastName"
date: today
format:
  html:
    toc: true
    toc-depth: 4
    code-fold: false
    theme: cosmo
    number-sections: true
    embed-resources: true
    include-in-header:
      text: |
        <style>
        kbd {
          display: inline-block;
          padding: 2px 6px;
          font-size: 0.85em;
          font-family: monospace;
          background-color: #f5f5f5;
          border: 1px solid #ccc;
          border-bottom: 2px solid #aaa;
          border-radius: 3px;
        }
        </style>
execute:
  echo: true
  warning: false
  message: false
  error: true
urlcolor: blue
linkcolor: blue
engine: knitr
---

```{r setup, include=FALSE, cache=FALSE, message=FALSE, warning=FALSE, results='hide'}
options(digits = 5)
knitr::opts_chunk$set(
  echo     = TRUE,
  cache    = FALSE,
  warning  = FALSE,
  message  = FALSE,
  collapse = FALSE,
  comment  = "",
  error    = TRUE    # never halt rendering on bash exit code 1
)

# ── Unset stale virtual environment from other projects ───────────────────────
Sys.unsetenv("VIRTUAL_ENV")

# ── Fix PATH for bash chunks on macOS (safe no-op on Linux/WSL2) ─────────────
# RStudio launched from the Dock inherits a minimal launchd PATH that omits
# /opt/homebrew/bin. This prepends Homebrew paths so knitr bash chunks see
# bash 5 and system tools like check_bash.
homebrew_paths <- c("/opt/homebrew/bin")
existing <- homebrew_paths[dir.exists(homebrew_paths)]
if (length(existing) > 0) {
  old_path <- Sys.getenv("PATH")
  new_path <- paste(c(existing, old_path), collapse = ":")
  Sys.setenv(PATH = new_path)
}
```

---

## Preliminaries

### Instructions for This Notebook

Follow instructor demo, render and submit at the end of lecture.

::: {.callout-tip}
### Command-line habits for this class

- **Tab-complete** directory and file names in the terminal. If nothing
  completes, you have a typo or are in the wrong directory.
- **Copy-paste** paths from this notebook rather than typing them.
  WSL2 Windows Terminal: <kbd>Ctrl</kbd>+<kbd>Shift</kbd>+<kbd>C</kbd> /
  <kbd>Ctrl</kbd>+<kbd>Shift</kbd>+<kbd>V</kbd>.
  macOS: <kbd>Cmd</kbd>+<kbd>C</kbd> / <kbd>Cmd</kbd>+<kbd>V</kbd>.
- **Develop in the terminal, memorialize in the notebook.** Write and test
  commands in your shell first. Once they work, paste them into the chunk.
- **Read errors before asking for help** — they almost always name the
  problem exactly.
- **In UNIX, silence is golden.** No output + exit code 0 = success.
:::

::: {.callout-note}
### Role Responsibilities

- **All Team Members** work individually to follow and reproduce the demo.
- **All Team Members** signal the Reporter or Scrum Master when blocked.
- **Scrum Master** leads the stand-up (~2 min), timeboxes work, signals
  instructor when team is ready or needs help.
- **Reporter** records blockages and resolutions in the log tables below,
  and summarizes the retrospective at the end.
- **Developers** test and troubleshoot, help blocked teammates.
:::

---

## Initial Steps Check and Render Guard

This chunk runs on every render. It never halts rendering — instead it
prints `✓` or `✗ INCORRECT` with a clear fix message for each check.
Read it carefully before proceeding with any step.

```{bash render-guard, error=TRUE}
echo "=== Render Guard ==="
echo ""

# ── 1. Working directory ──────────────────────────────────────────────────────
EXPECTED="$HOME/DSC-011-Sp26/DSC-011-S26_Lecture31_Demo_UNIX_Text_Utilities_II"
ACTUAL=$(pwd)
if [[ "$ACTUAL" == "$EXPECTED" ]]; then
    echo "✓ Working directory: $ACTUAL"
else
    echo "✗ INCORRECT — Wrong working directory"
    echo "  Rendering from: $ACTUAL"
    echo "  Expected:       $EXPECTED"
    echo "  Fix:"
    echo "    1. Close this project (File → Close Project)"
    echo "    2. Change directory to ~/DSC-011-Sp26/"    
    echo "    3. git clone <demo-URL>" 
    echo "    4. Open ~/DSC-011-Sp26/DSC-011-S26_Lecture31_Demo_UNIX_Text_Utilities_II/DSC-011-S26_Lecture31_Demo_UNIX_Text_Utilities_II.Rproj"
    echo "    5. Re-render"
fi

echo ""

# ── 2. bash version ───────────────────────────────────────────────────────────
BASH_MAJOR="${BASH_VERSINFO[0]}"
BASH_MINOR="${BASH_VERSINFO[1]}"
if [[ "$BASH_MAJOR" -gt 5 ]] || \
   { [[ "$BASH_MAJOR" -eq 5 ]] && [[ "$BASH_MINOR" -ge 1 ]]; }; then
    echo "✓ bash $BASH_VERSION at $(which bash)"
else
    echo "✗ INCORRECT — bash $BASH_VERSION is too old (need 5.1+)"
    echo "  macOS fix:"
    echo "    brew install bash"
    echo "    echo /opt/homebrew/bin | sudo tee /etc/paths.d/homebrew"
    echo "    Quit RStudio fully; relaunch: open -a RStudio"
    echo "  WSL2 fix:"
    echo "    sudo apt update && sudo apt install --only-upgrade bash"
fi

echo ""

# ── 3. check_bash ─────────────────────────────────────────────────────────────
if command -v check_bash &>/dev/null; then
    echo "✓ $(check_bash --version) at $(which check_bash)"
else
    echo "✗ INCORRECT — check_bash not found on PATH"
    echo "  Fix: complete Step 3 (clone repo, just install-system)"
    echo "  then re-render."
fi

echo ""

# ── 4. git available ──────────────────────────────────────────────────────────
if command -v git &>/dev/null; then
    echo "✓ $(git --version)"
else
    echo "✗ INCORRECT — git not found"
    echo "  macOS: brew install git"
    echo "  WSL2:  sudo apt install git"
fi

echo ""
echo "=== End Render Guard ==="
```

---

# LECTURE 30 DEMO: UNIX Text Utilities

## `grep`: Print all lines containing "ch", "Ch", "cH", or "CH" in the text file `data/processed/names2000s.txt` containing data on top baby names of the 2000s.

```{bash grep1, error=TRUE}
MY_CODE='grep -i ch data/processed/names2000s.txt'
eval "$MY_CODE" | 
  check_bash --code "$MY_CODE"   \
             --requires grep     \
             --requires-flag -i  \
             "544a26a357cb29a63612d11edfd13307ba5f049455007ed538c0b6295e9661e8"
```

## `cut -c`: Print characters of all lines from 5-15 and 25 in the text file `data/processed/names2000s.txt` containing data on top baby names of the 2000s.

```{bash cut1, error=TRUE}
MY_CODE='cut -c 5-15,25- data/processed/names2000s.txt'
eval "$MY_CODE" | 
  check_bash --code "$MY_CODE"   \
             --requires cut      \
             --requires-flag -c  \
             "62f4e4064bc52b0037f9a505600232214a5ab691cb3bdf3d26a87916f16f83ab"

```

## `cut -f`: Print fields 1,2 and 4 of every line using pipes (`|`) as delimiters in the text file `data/processed/names2000s.txt` containing data on top baby names of the 2000s.

```{bash cut2, error=TRUE}
MY_CODE='cut -d\| -f1,2,4 data/processed/names2000s.txt'
eval "$MY_CODE" | 
  check_bash --code "$MY_CODE"   \
             --requires cut      \
             --requires-flag -f  \
             --requires-flag -d  "e3b0c44298fc1c149afbf4c8996fb92427ae41e4649b934ca495991b7852b855"
```


## Shell Globs

Write a shell-glob to exactly match the files `data/processed/names1900s.txt`,`data/processed/names1910s.txt`,`data/processed/names2000s.txt`, and `data/processed/names2010s.txt` but no other of the processed data files. 

```{bash glob, error=TRUE}

MY_CODE=''
eval "$MY_CODE" | 
  check_bash --code "$MY_CODE"   \
             --requires ls      \
             --requires-flag -1  \
             "c0b7f82de96d3eb694e5c74dd93df7eb33e7b3d2e290e661d26ee32263b55308"

```


## Pipe and Filter Programming with `grep`

Write a shell-glob to **match all lines in any of the processed data across all 14 decades** that contain both a "z" and a "y" of either case.


```{bash pipe, error=TRUE}
MY_CODE=''
eval "$MY_CODE" | 
  check_bash --code "$MY_CODE"   \
             --requires grep     \
             --requires grep     \
             --requires-flag -i  \
             "243f00b955294c6871e45191880d3a8553ca56101a7f631aa144062773a41cb5"

```

### Answer These Questions:

1. Does this command pipeline match **names** containing **both** a "z" or "y" of any case, or **lines** containing **both** a "z" or "y" of any case, **or neither?** **Answer below in line 221** by deleting two of three of the following: **names lines neither**

**YOUR ANSWER: names lines neither**

2. Which two names in the output contain both a "z" and a "y"? 

**YOUR ANSWER: **

## Scrubbing Data to tab-separated values (TSV). 

Memorialize the pipeline that matches the line containing "Michael" but not "Michaela" in the 1990s data textfile and scrubs it to **tab-separated values format.**

```{bash scrub, error=TRUE}
MY_CODE=''
eval "$MY_CODE" | 
  check_bash --code "$MY_CODE"   \
             --requires grep     \
             --requires tr       \
             --requires tr       \
             "9da95e3fb15392a788f2dae4721c76fc867a4b5dd7c6093d161e2db8dbb83432"

```

## Extracting fields based on values in other fieldsl; finding unique and duplicated lines with `uniq` 

Which unique girls names had the same rank as "Michael" in any decade?

```{bash uniq, error=TRUE}
MY_CODE=''
eval "$MY_CODE" | 
  check_bash --code "$MY_CODE"   \
             --requires grep     \
             --requires cut       \
             --requires tr       \
             --requires sort       \
             --requires uniq       \
             "c737a32661e8dd70fb5b9cf5a1105133e28688909dca7912c7e8551878ce6027"

```

## Interactive Data Science in the UNIX Shell

Use `w3m` to interactively convert all of the raw HTML files with a shell glob, select datalines with `grep` and then use `tr`, `cut`, `sort` and `uniq` to answer: How many unique boys and girls names over all decades appear in the data? 

```{bash count-unique-names, error=TRUE}
MY_CODE=''
eval "$MY_CODE" | 
  check_bash --code "$MY_CODE"   \
             --requires w3m      \
             --requires grep     \
             --requires tr       \
             --requires tr       \
             --requires tr       \
             --requires tr       \
             --requires cut       \
             --requires sort       \
             --requires uniq       \
             --requires wc       \
             "9f150d56c8111d8236a69a2cfa52cc546964375087648bfb46cf2d5eb8625ec8"

```

2. In your terminal, build-out test and run each successive step in the pipeline from Jeroen Jannsons *Data Science at the Command Line* demonstrated in lecture and shown in the lecture slides, that streams the text of *Alice's Adventures in Wonderland*, converts all words to lower-case, matches and prints every word on a new line,  sorts the words ASCIIbetically (so duplicate lines are findable by `uniq`), counts the frequency of each word, sorts the words inversely by their frequency and prints the top 10 words


. 

```{bash coutnt-all-names, error=TRUE}
code=$()
check_bash "$code" "daae24db8bee2aaf3c4f0422311d887d1221aef0bd11b0068fb4a977d8381b0a"


```

---

## Reporter's Mini-Retrospective Summary

**What helped the team succeed today:**

*(write here)*

**What could make the next session run better:**

*(write here)*

---

::: {.callout-important}
## ★ LECTURE 31 DEMO 

**Render this notebook and submit the HTML to CatCourses.**
Before submitting, confirm:

- [ ] Your name is in the `author:` field
- [ ] All steps show `✓ CORRECT` 

Partial credit is awarded for notebooks with some `✗ INCORRECT` outputs.
:::

---

```{r session-info, echo=FALSE}
sessionInfo()
```
